---
title: "Simba4bioc: interfacing SIMBA's graph embedding methods to Bioconductor"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Simba4bioc: interfacing SIMBA's graph embedding methods to Bioconductor}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

SIMBA ([paper link](https://www.nature.com/articles/s41592-023-01899-8)) ([github](https://github.com/pinellolab/simba)) provides methods for "graph embeddings" of relationships between
elements assayed in multimodal single cell experiments.  The main
software stack is written in Python.  This package uses basilisk
to simplify use of the SIMBA methods in R/Bioconductor.

# Motivating illustration

The key component of the package is the "pinned"
simba python module.  Most of what is computed
in using SIMBA graph embeddings is defined
in this module.

```{r lk1, message=FALSE}
library(Simba4Bioc)
sref = simba_ref()
names(sref)
```

## Acquiring demonstration data

The "3k PBMC" dataset from TENx is a common
resource for demonstrations.  We have a compressed
version of it and the `get_` function will
uncompress and return the path of the h5ad
(AnnData) representation.  We then
use the `read_h5ad` method from simba
to access the dataset.

```{r dodec}
p3k = get_10x3kpbmc_path(overwrite=TRUE)
pp = sref$read_h5ad(p3k)
```
`pp` is now a proper AnnData instance.
```{r lkpp}
pp
names(pp)
pp['X'][1:3,1:3]
```

## Building and exploring the graph embedding

### Code and output

This sequence of commands will
filter genes and build a graph of
genes and cells as nodes and weighted edges
defined by discretized expression measures.
(See `tools/_general.py` in the simba
modules for details on the discretization
procedure.)

```{r dog1, eval=FALSE}
bb = basic_preproc(pp, simba_ref=sref)
gg = build_and_train_pbg( bb, simba_ref=sref )
dict = sref$read_embedding()
```

`dict` is list containing a pair of AnnData instances named
'C' and 'G' corresponding to cell and gene embeddings
respectively.  

### Embedding layout

Before we start exploring the embeddings numerically,
let's see what kinds of data elements are involved.
SIMBA uses the PyTorch BigGraph framework.  
For the operations noted above, the graph data
take the following form:
```
pbg
└── graph0
    ├── entity_alias.txt
    ├── graph_stats.json
    ├── input
    │   ├── edge
    │   │   └── edges_0_0.h5
    │   └── entity
    │       ├── entity_count_C_0.txt
    │       ├── entity_count_G_0.txt
    │       ├── entity_names_C_0.json
    │       └── entity_names_G_0.json
    ├── model
    │   ├── checkpoint_version.txt
    │   ├── config.json
    │   ├── embeddings_C_0.v10.h5
    │   ├── embeddings_G_0.v10.h5
    │   ├── model.v10.h5
    │   └── training_stats.json
    └── pbg_graph.txt
```
The folder names `graph0` and `model` are arbitrary, set as
defaults in the `build_and_train_pbg` function.
The `graph0/model/embeddings...` HDF5 files hold
the coordinates of cells and genes in the graph-derived
latent spaces.  Thus an invocation of `rhdf5::h5read` on
the `embeddings_C_0.v10.h5` file will produce a 50 x 2700
matrix of FLOAT.

### Figure 2b: UMAP for cell embedding

To reproduce Figure 2b of the SIMBA paper
without recomputing the graph embedding
we have stored the cell embedding in the package,
for retrieval using `read_h5ad(get_3k_cell_emb())`.
The cell type labels in the initial AnnData instance `pp`
computed above need to be reorganized.

```{r dofig2b, message=FALSE}
cemb = sref$read_h5ad(get_3k_cell_emb())
cemb$obs$tmp = pp[cemb$obs_names]$obs[["celltype"]]
names(cemb$obs) = "celltype"
library(uwot)
set.seed(1234)
cumap = umap(cemb['X'])
library(ggplot2)
mydf = data.frame(x=cumap[,1], y=cumap[,2], 
   ctype=cemb$obs$celltype)
ggplot(mydf, aes(x=x,y=y,colour=ctype)) + geom_point()
```

### Figure 2c: Embedding genes and cells together

We won't produce figure 2c exactly, as 2c appears
to use gene filtering.  We saved the gene
embedding in `gemb` to avoid building the
graph repeatedly in this vignette.

```{r doemb, message=FALSE}
gemb = sref$read_h5ad(get_3k_gene_emb())
ginc = sref$tl$embed(adata_ref=cemb, list_adata_query=list(gemb))
eanno = c(as.character(cemb$obs$celltype), rep("gene", nrow(gemb['X'])))
fullanno = rownames(ginc$obs)  # cell types and gene symbols concatenated
ginc$obs$entity_anno = eanno
```

Now `ginc` is an AnnData instance with 'X' component the
50-dimensional latent space constructed from the graph
linking genes and cells by expression-based weights.
UMAP is applied to present the 2-d reduction:

```{r dopl, message=FALSE}
library(uwot)
uuu = umap(ginc['X'])
kkdf = data.frame(x=uuu[,1], y=uuu[,2], 
   ent=ginc$obs$entity_anno, fullanno=fullanno)
ggplot(kkdf, aes(x=x, y=y, colour=ent, text=fullanno)) + 
   geom_point(size=.6, alpha=.5)
```

When this plot is rendered interactively with plotly, points
corresponding to genes are annotated with gene symbol.
To see how this works in a vignette, we'll drill
down on a small portion of the projection.

To help with annotation, we define a new function
that adds a persistent label.
```{r drill, message=FALSE}
library(plotly)
addg = function (p=., sym = "NKG7", dat) 
{
    ro = dplyr::filter(dat, fullanno == sym)
    plotly::add_text(p, x = ro$x, y = ro$y, text = sym, 
        textfont=list(size=15, color="black"))
}
```

Now produce the interactive visualization with selected
annotations of gene positions in the combined embedding.

```{r annopl}
ndf = kkdf |> dplyr::filter(x > -6.2, x < -3, y > 4.5)
p <- plot_ly(ndf, x=~x, y=~y, color=~ent,
   mode="markers", text=~fullanno, type="scatter") %>% 
     addg(dat=ndf) %>% 
     addg(sym="IL7R", dat=ndf) %>%
     addg(sym="CD8B", dat=ndf) %>%
     addg(sym="CD8A", dat=ndf) %>%
     addg(sym="BCL2", dat=ndf) %>%
     addg(sym="LDLRAP1", dat=ndf)
p
```

### Figure 2d: barcodes

We'll reuse the `pp` representation of the 3k PBMC
data computed above to illustrate the barcode
plotting for multiple genes.

```{r do4, message=FALSE}
syms = c("CST3", "MS4A1", "NKG7", "GAPDH")
cp = get_3k_cell_emb()
gp = get_3k_gene_emb()
dats = lapply(syms, function(x) {
  ans = simba_barcode_data(symbol = x, origad = pp,
    clabel="celltype", simbaref=sref, cembpath = cp,
    gembpath = gp)
  ans$gene = x
  ans
  })
alld = do.call(rbind, dats)
library(ggplot2)
ggplot(alld, aes(x=rank, xend=rank, y=0, yend=score, colour=attr)) +
  geom_segment() + facet_wrap(~gene, ncol=2) + 
    theme(strip.text=element_text(size=24), 
      legend.text=element_text(size=24), 
      axis.text=element_text(size=24))
```

# Conclusions

This takes care of the bulk of Figure 2 of the initial SIMBA
paper. Data for 2e can be obtained from 
```
Simba4Bioc:::compare_entities(sref, cemb, gemb)
```

Additional vignettes in production will explore multiomic applications.
